Capon Beamforming Applied on Moving Objects - An Analysis of Local Lateral Shift-Invariance

### 2013-11-22 HRUIofMovingObjects/jrnl_article.tex
Moving object article: done some writing. Added some figures

An important property of an ultrasound imaging system is its local shift-invariance. Local shift-invariance means that an object's appearance should be preserved when the object, or the ultrasound probe, is moved slightly. This invariance for local movements is what makes ultrasound imaging feasible in general, but it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood. In this paper we analyze the lateral shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects. We also show some \textit{in-vivo} examples.

As mentioned, Capon beamforming has been shown to provide improved lateral resolution and edge definitions in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}. The method accomplishes this by adjusting the array weights based on element-to-element covariance. Hence, the method yields data dependent weights. Capon beamforming is however also known for its suboptimal performance if the sample size available for covariance estimation is limited \cite{Mestre2006}, if correlated signals are present \cite{Widrow1982}, and if the assumed steering vector does not match the signal propagation vector \cite{Wax1996, Wax1996a}. This has led to the development of several robust versions of the Capon beamformer with e.g. spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying steering vector uncertainty sets (SVUS) \cite{Lorenz2005, Rubsamen2013}. The first two, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}. The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics \cite{Synnevag2007a}. SVUS reduces the signal-nulling effect of the Capon beamformer which appears if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly. This self-nulling is actually how the method obtains its super resolution capabilities, but the signal can be lost in the process. In this paper we make use of a steering vector set to reduce this self-nulling effect.

In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer, and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations. Cox states the following important fact about the Capon beamformer (the $\vec{k}_3$ processor): 
\begin{quote}
"\textit{... a $\vec{k}_3$ processor requires more closely spaced beams than a ... $\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}", 
\end{quote}
where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer. In Fig.\,\ref{fig:das_capon_beams} we have plotted the steered response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}. A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution. Impinging on the array is a monochromatic plane wave with a propagation vector orthogonal to the array surface with no spatially white noise present. Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity. The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output. A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 20 dB before array gain is applied. The figure shows that when the DAS beamformer is used with a beamspacing equal to the system resolution, only a 3.92 dB scalloping loss \cite{Harris1978} should be observed if the source of the plane wave starts moving parallel to the array. However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $31$ dB.

@@

In previous literature on Capon beamforming \comment{cite earlier work} for medical ultrasound imaging (with single frame simulations) this effect has been passed over in silence by either oversampling on transmit, or by carefully position point scatterers exactly on transmit-receive beam pairs, or by ignorance. In our first work where Capon beamforming was applied on a loop of images containing moving objects \cite{Asen2012, Asen} we commented on the effect, but no solution was presented other than oversampling on transmit when simulating.

In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS the predicted scalloping loss is observed, manifested as oscillating point scatteres. The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate. In this paper we therefore study the scalloping loss of the Capon beamforming and investigate different methods aiming at reducing this effect, and how the different methods will affect the imaging frame rate and the processing complexity.

In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and shift invariance. In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before the different methods are presented in Section \ref{sec:methods}. Then, in Section \ref{sec:res} and Section \ref{sec:dis} we present and discuss the results of applying the most promising method on \textit{in-vitro} and \textit{in-vivo} data . Finally we give our conclusion in Section \ref{sec:con}.

@@

The sample covariance matrix can be estimated for an active system in the following way \cite{Synnevag2009}:
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,
\end{align}
where $K = M-L+1$ typically is proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is the $l\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$ of length $L$. Finally the matrix is loaded with a diagonal factor for numerical stability and increased robustness, 
\begin{align}
\epsilon &= d*\text{trace}\{\mat{\breve{R}}\}/L\\
\mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
\end{align}

@@

Note that one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system. This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}. From (\ref{eq:capon_weights}) it is clear that complex data will result in complex weights. The weights can therefore have an asymmetric K-space response which basically means that zeros does not have to be symmetric around the origin and that the beam can be micro-steered by phase rotation.

@@

The lateral sampling required in an ultrasound image is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$), and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the body and $\lambda_c$ is the wavelength of the center frequency. The required Nyquist beam spacing is then \cite{Hergum2007}:
\begin{align}
\Delta < \frac{\lambda_c}{D_{tx} + D_{rx}}. \label{eq:resolution}
\end{align}
If this requirement is not fulfilled the scalloping loss could become visible in the image and the system will no longer be lateral shift-invariant. In an ultrasound image with 256 gray levels and 50 dB dynamic range a 1 dB loss corresponds to 5 gray levels. This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1dB could therefore end up being visible to the observer. In our analysis a system would, based on this reasoning, be regarded as locally lateral shift-invariant if the beam-too-beam loss is smaller than 1 dB. However, depending on light conditions and e.g. the selected gray level mapping, a larger loss might be acceptable.

In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}. A LSV-plot is constructed for a given imaging system by imaging a laterally moving point scatterer. For each lateral position the root-mean-square (RMS) is calculated in range, and the resulting beam profiles are stacked to produce an image. The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB. The maximum for each beam profile is also marked with a black dot. The system is said to be local lateral shift-invariant if the contours are diagonal, hence the diagonals have a loss less than 1 dB.

To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point in each beam profile versus $f(x)=x$ (P-MAE), and the MAE of the amplitude variation for the same set of points (A-MAE). The first number quantifies the amount of geometric distortion, and the latter quantifies the gain variation or in our case the loss of signal. The maximum amplitude deviation along the max-point trace is also calculated and reported. We present all these numbers together with the LSV-plots.

Fig.\,\ref{fig:das_shiftinvariant} presents a LSV-plot for an imaging system using DAS beamforming with a beam spacing equal to (\ref{eq:resolution}). The red dotted vertical lines indicates the position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} \comment{add table with simulation settings or just insert in text?}. We see that the contours are approximately diagonal and therefore we can conclude that the system is locally lateral shift-invariant. The P-MAE and A-MAE are also minimal.

@@

In Fig.\,\ref{fig:das_shiftvariant} the beam spacing is reduced by a factor of two. We see how the scalloping loss is increased to a level where the contours no longer are diagonal. Hence, the imaging system is no longer locally lateral shift-invariant. We observe that the maximum peak is lost when the point is located between two beams and the distance between the beams are too high.

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
\hfill{}
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_undersampled2_2x.eps}\label{fig:das_shiftvariant}}
}
\caption{LSV-plots of DAS beamforming. If the contours are diagonal the imaging system is said to be lateral local shif-invariant. Red vertical lines are plotted where transmit-receive beam pairs are located. a) Sampling corresponding to (\ref{eq:resolution}). The system is locally lateral shift-invariant. b) Undersampling with a factor two. The system is no longer locally lateral shift-invariant.}
\label{fig:das}
\end{figure*}

\section{Local Lateral Shift-Invariance of the Capon Beamformer}\label{sec:capon_LLSI}
We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the once used in earlier work \cite{Synnevag2009, Asen}. In Fig.\,\ref{fig:capon1}. the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 32$, $K=1$ and $d=1/100$. The set of lateral displacements is constructed in such a way that the center beam is the only one hitting the point scatterer exactly. We see that the scalloping loss is tremendous, and not far from the value predicted by Fig\,\ref{fig:das_capon_beams}. The difference between the predicted and the simulated loss might be caused by several factors. The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS used in Fig.\,\ref{fig:capon1} average a broadband signal in range. For a slice through the center of the pulse the scalloping loss is below -30dB.

The diagonal factor $d$ and the effective adaptive aperture size $L$ can be used to control the mainlobe width of the Capon beamformer. In both extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes the DAS beamformer, and will require a beamspacing according to (\ref{eq:resolution}). In Fig.\,\ref{fig:capon2} we show how the scalloping loss is reduced when the amount of diagonal loading is increased. We observe that Fig.\,\ref{fig:capon2} contains the same pattern as present in Fig.\,\ref{fig:das_shiftvariant}. From Fig.\,\ref{fig:capon} it should be obvious that the Capon beamformer either require increased lateral sampling in order to make the imaging system locally lateral shift-invariant, or the resolution has to be reduced to the same level as DAS. It would be unwise to do all the computation involved with Capon beamforming with a high $d$ value and end up with a resolution equal to DAS. However, we see that the diagonal factor can be a useful tool if resolution has to be traded for the need of oversampling. The same is true for the subarray length parameter.

\begin{figure*}[!t]
\centerline{
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2.eps}\label{fig:capon1}
}
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=1_2.eps}\label{fig:capon2}
}
}
\caption{LSV-plots of Capon beamforming with $L=M/2=32$ and $K=1$, and with sampling corresponding to (\ref{eq:resolution}) a) The diagonal loading factor $d=1/100$. The system is not locally lateral shift-invariant. b) The diagonal loading factor $d$ is increased to 1. The system is still not locally lateral shift-invariant.}
\label{fig:capon}
\end{figure*}

\section{Methods}\label{sec:methods}
We will now discuss different ways to improve on the lateral shift-invariance when the Capon beamformer is used. From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally lateral shift-invariant. In this section we will investigate how this oversampling should be conducted, how large it needs to be and how the proposed methods will affect the system frame rate and the processing complexity.

\subsection{Oversampling on Transmit}
The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until (\ref{eq:resolution}) is fulfilled or the LSV-plot has an uniform level along the diagonals. In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten respectively. As expected the scalloping loss decreases with increased lateral sampling. Note that Fig.\,\ref{fig:capon_oversampling_b} is zoomed compared to Fig.\,\ref{fig:capon} in order to better see the beam-to-beam variation.

From Fig.\,\ref{fig:capon_oversampling_b} we see that more than ten times oversampling has to be applied to achieve local lateral shift-invariance when a moving point scatterer is imaged without noise and with the same parameters as in Fig\,\ref{fig:capon}. %$1/100$ in diagonal loading. 
Note that a diagonal loading factor of $1/100$ will introduce a deterministic white noise component in the sample covariance matrix $40$ dB lower than the signal present. Hence, the covariance has been estimated from data which appeared to have an SNR of 40 dB. \comment{correct?}

Even though increasing the number of transmits improves the lateral shift-invariance it will also reduce the imaging frame rate with a factor equal to the oversampling factor. The processing complexity will also increase equally. For these two reasons, oversampling on transmit is not a feasible option for real time ultrasound imaging, and we will not discuss this method further.

@@

\subsection{Oversampling by Parallel Receive Beamforming}
Parallel receive beamforming (PRB) means that multiple narrow receive beams (rx-beams) are constructed from one broad tx-beam. Then, by reducing the number of tx-beams and relying on multiple rx-beams in order to get sufficient sampling on receive, it is possible to achieve high frame rates in e.g. cardiac ultrasound imaging. However, PRB also introduce motion artifacts in the image which to some degree can be compensated for \cite{Hergum2007} \comment{cite Bastien}. In (\comment{Cite paper where they use STB + Capon if its out}) it is shown that synthetic transmit beamforming (STB) \cite{Hergum2007} works with Capon beamforming. In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording. In the following discussion we consider PRB as being equal to oversampling on transmit when it comes to image quality.

In cardiac ultrasound imaging, PRB is typically used in order to keep the number of rx-beams while the number of tx-beams are reduced. When using PRB for oversampling we would like to keep the number of tx-beams while the number of rx-beams are increased. The acquisition frame rate will then not be affected. On the other hand more rx-beams will increase the number of covariance matrices, and thereby the number of matrix-inversions with a factor equal to the number of parallel receive lines. The processing times reported in \cite{Asen} will then also increase with the same factor. Since almost 10x oversampling is required in order to achieve lateral shift invariant imaging of a moving point scatterer, non of the implementations described in \cite{Asen} would be real time any more.

One way of reducing the number of matrix inversions is to estimate the covariance matrix from multiple beams \cite{Jensen2012}, generating one covariance matrix per range.  Tailoring the method of Jensen and Austeng to cardiac ultrasound imaging it makes more sense, due to motion, to construct one covariance matrix per tx-beam using multiple parallel rx-beams. With this solution the number of matrix inversions will not changes, but the time taken to construct the matrices will increase. We will not investigate this method further in this paper. \comment{Kanskje litt dumt å nevne det da...?} \comment{move to discussion}

\subsection{Oversampling by Phase Rotation}
The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation. The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since the ultrasound data are pre-delayed. However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors. In a narrow band setting the beam can make a sweep across the whole sector by only using phase rotation. This is also possible in broadband applications as long as the phase rotation is limited to a fraction of the wave length \comment{cite fine-coarse beamformer}. It is therefore possible to make a sweep between the current beam and its nearest neighbours using only phase rotation. Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.

For an uniform linear array located along the x-axis the steering vector $\vec{a}$ should be calculated as 
\begin{align}
\vec{a} = 
\begin{bmatrix}
e^{-j\frac{2\pi}{\lambda_c}x_0\sin(\theta)} \\
e^{-j\frac{2\pi}{\lambda_c}x_1\sin(\theta)} \\
\vdots \\ 
e^{-j\frac{2\pi}{\lambda_c}x_{L-1}\sin(\theta)}
\end{bmatrix}
\end{align}
where $x_i$ is the element position and $\theta$ is swiped from $-\Delta/2$ too $\Delta/2$ where $\Delta$ is as defined in (\ref{eq:resolution}). Note that the steering vector is calculated for a $L$ element array centered around the origin, even though the full array has $M$ elements. It is important that this synthetic array has the same element pitch as the full array.

@@

\subsection{In-vitro Images}
Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally. Confirm the improved shift-invariance when using phase rotation.

>>>>

An important property of an ultrasound imaging system is its local shift-invariance. Local shift-invariance means that an object's appearance should be preserved when the object, or the ultrasound probe, is moved slightly. This invariance for local movements is what makes ultrasound imaging feasible in general, but it is specially important for quantitative techniques that correlate time-varying features, like pulsed doppler and flow imaging, and speckle tracking of tissue and blood. In this paper we analyze the lateral shift-invariance property of the Capon beamformer by studying the beamformer output when applied on simulated and \textit{in vitro} images of moving objects.

As mentioned, Capon beamforming has been shown to provide improved lateral resolution and edge definitions in ultrasound images \cite{Synnevag2007, Synnevag2009, Chen2011}. The method accomplishes this by adjusting the array weights based on element-to-element covariance. Hence, the method yields data dependent weights. Capon beamforming is however also known for its suboptimal performance if the sample size available for covariance estimation is limited \cite{Mestre2006}, if correlated signals are present \cite{Widrow1982}, and if the assumed steering vector does not match the signal propagation vector \cite{Wax1996, Wax1996a}. This has led to the development of several robust versions of the Capon beamformer with e.g. spatial smoothing \cite{Shan1985} or diagonal loading \cite{JianLi2003} of the covariance estimate, or by applying steering vector uncertainty sets (SVUS) \cite{Lorenz2005, Rubsamen2013}. The first two, together with temporal smoothing of the covariance estimate, have been applied in earlier work on Capon beamforming for medical ultrasound imaging \cite{Synnevag2009}. The first two trade resolution for robustness and temporal smoothing helps to preserve speckle statistics \cite{Synnevag2007a}. SVUS reduces the signal-nulling effect of the Capon beamformer which appears if the signal is present during the covariance estimation process and if this estimate is later used with a steering vector that does not match the signal propagation vector exactly. This self-nulling is actually how the method obtains its super resolution capabilities, but the signal can also get lost in the process. In this paper we make use of a steering vector set to reduce this self-nulling effect.

In \cite{Cox1973}, Cox derives formulas for the resolution of the Capon beamformer, and presents in great detail how the amount of mismatch between the selected steering vector and the signal propagation vector impacts the beamformer output in simple situations. Cox states the following important fact about the Capon beamformer (the $\vec{k}_3$ processor): 
\begin{quote}
"\textit{... a $\vec{k}_3$ processor requires more closely spaced beams than a ... $\vec{k}_1$-processor in order to avoid serious signal suppression effects being introduced on signals arriving from directions between the beams}", 
\end{quote}
where $\vec{k}_1$ is the delay-and-sum (DAS) beamformer. In Fig.\,\ref{fig:das_capon_beams} we have plotted the steered response for DAS and the Capon beamformer using (26) and (32) in \cite{Cox1973}. A 64-element linear array with $\lambda/2$ element spacing, where $\lambda$ is the wavelength, is scanned over a sector equal to three times the system resolution ($1/32$ rad). Impinging on the array is a monochromatic plane wave with a propagation vector orthogonal to the array surface with no spatially white noise present. Hence, the interference-noise matrix ($\sigma_0^2\mat{Q}$ in \cite{Cox1973}) is the identity. The effect of typical subarray averaging and diagonal loading ($L=32$ and $d=1/100$) has also been added to the Capon beamformer output. A diagonal loading of $1/100$ corresponds to an signal-to-noise ratio of 20 dB before array gain is applied. The figure shows that when the DAS beamformer is used with a beamspacing equal to the system resolution, only a 3.92 dB scalloping loss \cite{Harris1978} should be observed if the source of the plane wave starts moving parallel to the array. However if the Capon beamformer is used with the same beam density the scalloping loss could be as large as $31$ dB.

@@

In previous literature on Capon beamforming for medical ultrasound imaging (with single frame simulations) this effect has been passed over in silence by either oversampling on transmit, or by carefully position point scatterers exactly on transmit-receive beam pairs, or by ignorance. In our first work where Capon beamforming was applied on a loop of images containing moving objects \cite{Asen2012, Asen} we commented on the effect, but no solution was presented other than oversampling on transmit when simulating. The same was done by Jensen and Austeng \cite{Jensen2012}.

In this paper we investigate the issue further, and we show that when a moving point scatterer is imaged with the Capon beamformer with the same beam spacing sufficient for DAS the predicted scalloping loss is observed, manifested as blinking point scatteres. The question is then, what can we do in order to improve on the situation without affecting the imaging frame rate. In this paper we therefore study the scalloping loss of the Capon beamforming and investigate different methods aiming at reducing this effect, and how the different methods will affect the imaging frame rate and the processing complexity.

In the next section we give an introduction to Capon beamforming for medical ultrasound imaging together with a short introduction to lateral sampling and shift invariance. In Section \ref{sec:capon_LLSI} the local lateral shift-invariance of the Capon beamformer is investigated, before the different methods are presented in Section \ref{sec:methods}. Then, in Section \ref{sec:res} and Section \ref{sec:dis} we present and discuss the results of applying the most promising method on \textit{in-vitro} and \textit{in-vivo} data. Finally we give our conclusion in Section \ref{sec:con}.

@@

The sample covariance matrix can be estimated for an active broadband system in the following way \cite{Synnevag2009}:
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,
\end{align}
where $K = M-L+1$ typically is proportional to the pulse length, $N_K = 2K + 1$, $N_L = M-L+1$, and $\vec{x}_l$ is the $l\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$ of length $L$. Finally the matrix is loaded with a diagonal factor for numerical stability and increased robustness, 
\begin{align}
\epsilon &= d*\text{trace}\{\mat{\breve{R}}\}/L\\
\mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
\end{align}

@@

Note that one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system. This is indeed computational demanding, but following the innovation in GPU computing, and by applying a novel method in order to reduce the size of the matrix inversion problem \cite{Nilsen2009}, real time processing is now possible \cite{Asen}. From (\ref{eq:capon_weights}) it is clear that complex data can lead to complex weights. The weights can therefore have an asymmetric K-space response which basically means that the main lobe and zeros does not have to be symmetric around the origin. The beam can therefore be micro-steered by phase rotation.

@@

The lateral sampling required in an ultrasound image is given by the combined aperture size on transmit ($D_{tx}$) and receive ($D_{rx}$), and the center frequency ($f_c = c/\lambda$), where $c$ is the speed of sound in the body and $\lambda_c$ is the wavelength of the center frequency. The required Nyquist beam spacing is then \cite{Hergum2007}:
\begin{align}
\Delta < \frac{\lambda_c}{D_{tx} + D_{rx}}. \label{eq:resolution}
\end{align}
If this requirement is not fulfilled the scalloping loss could become visible in the image and the system will no longer be locally lateral shift-invariant. In an ultrasound image with 256 gray levels and 50 dB dynamic range, a 1 dB loss corresponds to 5 gray levels. This is approximately equal to the visibility threshold (Weber fraction of 2\%) for a grayscale image, and a loss larger than 1dB could therefore end up being visible to the observer. In our analysis a system would, based on this reasoning, be regarded as locally lateral shift-invariant if the loss in between beams is smaller than 1 dB. However, depending on light conditions and e.g. the selected gray level mapping, a larger loss might be acceptable.

In order to visualize the local lateral shift-invariance of a given imaging system, we will make use of lateral shift-variance plots (LSV-plots), introduced by Hergum \textit{et al.} \cite{Hergum2007}. A LSV-plot is constructed for a given imaging system by imaging a laterally moving point scatterer. For each lateral position of the point scatterer the root-mean-square (RMS) is calculated in the range dimension, and the resulting beam profiles are stacked to produce an image. The LSV-plots in this paper are displayed as a contour plot with contours on -1, -2, -3, -6, -12 and -24 dB. The maximum for each beam profile is also marked with a black dot. The system is said to be locally lateral shift-invariant if the contours are diagonal, hence the diagonals have a loss less than 1 dB.

To further quantify the shift-invariance we have calculated the mean absolute error (MAE) of the maximum point in each beam profile versus a straight line ($f(x)=x$). We name this measure P-MAE. The MAE of the amplitudes for the same set of maximum points versus the maximum in the whole plot is also measured. We call this measure A-MAE. The first measure quantifies the amount of geometric distortion, and the latter quantifies the gain variation or in our case the loss of signal. The maximum amplitude deviation along the max-point trace, or the scalloping loss, is also calculated and reported. We present all these numbers together with the LSV-plots.

Figure \ref{fig:das_shiftinvariant} presents a LSV-plot for an imaging system using DAS beamforming with a beam spacing equal to (\ref{eq:resolution}). The red dotted vertical lines indicates the position of transmit-receive beam pairs. The point scatterer response has been simulated using Field II \cite{Jensen1992, Jensen1996a} with a probe setup mimicking the M5S probe by GE Vingmed Ultrasound, Horten, Norway. The probe was excited with a 2.5 MHz and x cycles long pulse \comment{Add correct number of cycles}. We see that the contours are approximately diagonal and therefore we can conclude that the system is locally lateral shift-invariant. The P-MAE and A-MAE are also minimal.

@@

In Fig.\,\ref{fig:das_shiftvariant} the beam density is reduced by a factor of two. We see how the scalloping loss is increased to a level where the contours no longer are diagonal. Hence, the imaging system is no longer locally lateral shift-invariant. We observe that the maximum peak is greatly reduced when the point is located between two beams and the distance between the beams are too high.

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_normal2.eps}\label{fig:das_shiftinvariant}}
\hfill{}
\subfloat[]{\includegraphics[width=0.5\linewidth]{\img das_undersampled2_2x.eps}\label{fig:das_shiftvariant}}
}
\caption{LSV-plots of delay-and-sum beamforming. If the contours are diagonal the imaging system is said to be locally lateral shift-invariant. Red vertical lines are plotted where transmit-receive beam pairs are located. a) Sampling corresponding to (\ref{eq:resolution}). The system is locally lateral shift-invariant. b) Undersampling with a factor of two. The system is no longer locally lateral shift-invariant.}
\label{fig:das}
\end{figure*}

\section{Local Lateral Shift-Invariance of the Capon Beamformer}\label{sec:capon_LLSI}
We will now investigate the local lateral shift-invariance of the Capon beamformer with parameters equal to the ones used in earlier work \cite{Synnevag2009, Asen}. In Fig.\,\ref{fig:capon1} the same point scatterer response as presented in Fig.\,\ref{fig:das_shiftinvariant} has been weighted with Capon weights calculated using (\ref{eq:capon_weights}) with $L = M/2 = 32$, $K=1$ and $d=1/100$. The set of lateral displacements is constructed in such a way that the center beam is the only one hitting the point scatterer exactly. We see that the scalloping loss is tremendous, and not far from the value predicted by Fig\,\ref{fig:das_capon_beams}. The difference between the predicted and the simulated loss might be caused by several factors. The most important one being that Fig.\,\ref{fig:das_capon_beams} shows a single frequency whereas the RMS operator used in Fig.\,\ref{fig:capon1} average a broadband signal in range. For a slice through the center of the pulse the scalloping loss is below -30dB.

The diagonal factor $d$ and the effective adaptive aperture size $L$ can be used to control the mainlobe width of the Capon beamformer. In both extreme cases, $d\gg1$ and $L=1$, the Capon beamformer becomes either a triangle or uniform-weighted DAS beamformer respectively, and will require a beamspacing according to (\ref{eq:resolution}). In Fig.\,\ref{fig:capon2} we show how the scalloping loss is reduced when the amount of diagonal loading is increased. We observe that Fig.\,\ref{fig:capon2} now contains a scaled version of the pattern presented in Fig.\,\ref{fig:das_shiftvariant}. From Fig.\,\ref{fig:capon} it should be obvious that the Capon beamformer either requires increased lateral sampling in order to make the imaging system locally lateral shift-invariant, or the resolution has to be reduced to the same level as DAS. It would be unwise to do all the computation involved with Capon beamforming with a high $d$ value and end up with a resolution equal to DAS. However, we see that the diagonal factor can be a useful tool if resolution has to be traded for the need of oversampling. The same is true for the subarray length parameter.

\begin{figure*}[!t]
\centerline{
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=001_2.eps}\label{fig:capon1}
}
\subfloat[]{
\includegraphics[width=0.5\linewidth]{\img capon_L=32_K=1_d=1_2.eps}\label{fig:capon2}
}
}
\caption{LSV-plots of Capon beamforming with $L=M/2=32$ and $K=1$, and with sampling corresponding to (\ref{eq:resolution}). a) Capon beamforming with $d=1/100$ as diagonal loading. The system is not locally lateral shift-invariant. b) The diagonal loading factor $d$ is increased to 1. The system is still not locally lateral shift-invariant.}
\label{fig:capon}
\end{figure*}

\section{Methods}\label{sec:methods}
We will now discuss different ways of improving the local lateral shift-invariance when the Capon beamformer is used. From the previous section it should be clear that we somehow need to increase the lateral sampling to make the Capon imaging system locally lateral shift-invariant. In this section we will investigate how this oversampling should be conducted, how large it needs to be and how the proposed methods will affect the system frame rate and the processing complexity.

\subsection{Oversampling on Transmit}
The straight forward way of obtaining sufficient sampling in an active system is by increasing the number of transmit beams (tx-beams) until the system is laterally local shift-invariant. In Fig.\,\ref{fig:capon_oversampling_a} and Fig.\,\ref{fig:capon_oversampling_b} we have increased the lateral sampling on transmit by a factor of four and ten respectively. As expected the scalloping loss decreases with increased lateral sampling. Note that Fig.\,\ref{fig:capon_oversampling_b} is zoomed compared to Fig.\,\ref{fig:capon} in order to better see the beam-to-beam variation.

From Fig.\,\ref{fig:capon_oversampling_b} we see that more than ten times oversampling has to be applied to achieve a local lateral shift-invariant Capon beamformer when a moving point scatterer is imaged without noise and with the same parameters as in Fig\,\ref{fig:capon}. %$1/100$ in diagonal loading. 
Note that a diagonal loading factor of $1/100$ will introduce a deterministic white noise component in the sample covariance matrix $20$ dB lower than the signal present. Hence, the covariance has been estimated from data which appeared to have an SNR of 20 dB.

Even though increasing the number of transmits improves the lateral shift-invariance it will also reduce the imaging frame rate with a factor equal to the oversampling factor. The processing complexity will also increase equally. For these two reasons, oversampling on transmit is something which is desirable to avoid in real time ultrasound imaging, and therefore we will not discuss this method further.

@@

\subsection{Oversampling by Parallel Receive Beamforming}
Parallel receive beamforming (PRB) means that multiple narrow receive beams (rx-beams) are constructed from one broad tx-beam. Then, by reducing the number of tx-beams and relying on multiple rx-beams in order to get sufficient sampling on receive, it is possible to achieve high frame rates in e.g. cardiac ultrasound imaging. However, PRB also introduces motion artifacts in the image which to some degree can be compensated for \cite{Hergum2007} \comment{cite Bastien}. In (\comment{Cite paper where they use STB + Capon if its out}) it is shown that synthetic transmit beamforming (STB) \cite{Hergum2007} works with Capon beamforming. In \cite{Asen} we also used STB in order to increase the beam density of a cardiac ultrasound recording. In the following discussion we consider PRB as being equal to oversampling on transmit when it comes to image quality, and the LSV-plots for PRB is therefore not included.

In cardiac ultrasound imaging, PRB is typically used in order to keep the number of rx-beams high while the number of tx-beams are reduced. When using PRB for oversampling we would like to maintain the same number of tx-beams while the number of rx-beams are increased. The acquisition frame rate will then not be affected. On the other hand more rx-beams will increase the number of covariance matrices, and thereby the number of matrix-inversions with a factor equal to the number of parallel receive lines. The processing times reported in \cite{Asen} will then also increase with the same factor. Since almost 10x oversampling is required in order to achieve lateral shift invariant imaging of a moving point scatterer, none of the implementations described in \cite{Asen} would be real time any more.

\subsection{Oversampling by Phase Rotation}
The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation. The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since the ultrasound data are pre-delayed. However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors. In a narrow band setting the beam can make a sweep across the whole sector by only using phase rotation. This is also possible in broadband applications as long as the phase rotation is limited to a fraction of the wave length (coarse-fine beamforming \cite{Thomenius}). It is therefore possible to make a sweep between the current beam and its two nearest neighbours using only phase rotation. Phase rotation is actually what the Capon beamformer use internally to micro-steer away from bright point scatterers and to enhance edges.

For an uniform linear array located along the x-axis the steering vector $\vec{a}$ should be calculated as 
\begin{align}
\vec{a} = 
\begin{bmatrix}
e^{-j\frac{2\pi}{\lambda_c}x_0\sin(\theta)} \\
e^{-j\frac{2\pi}{\lambda_c}x_1\sin(\theta)} \\
\vdots \\ 
e^{-j\frac{2\pi}{\lambda_c}x_{L-1}\sin(\theta)}
\end{bmatrix}
\end{align}
where $x_i$ is the element position and $\theta$ is swiped from $-\Delta/2$ to $\Delta/2$ where $\Delta$ is as defined in (\ref{eq:resolution}). Note that the steering vector is calculated for a $L$ element array centered around the origin, even though the full array has $M$ elements. It is important that this subarray has the same element pitch as the full array.

@@

%\subsection{In-vitro Images}
%Acquire channel data from a phantom containing point wire targets and cysts while moving the probe laterally. Confirm the improved shift-invariance when using phase rotation.
